---
title: "Template"
author: "Studentnames and studentnumbers here"
date: "`r Sys.Date()`"
output: pdf_document
---

# Set-up your environment

```{r package_install, include=FALSE}
install.packages("tidyverse")
install.packages("yaml")
install.packages("rmarkdown")
install.packages("eurostat")
install.packages("maps")
```

```{r packages}
require(tidyverse)
require(rmarkdown)
require(yaml)
require(eurostat)
require(maps)
```

# Title Page

Include your names: Imane Ouahi, David Abelovic, Xi Hui, Rebecca Devasia Let's do it!!! Hi Rebecca

Include the tutorial group number: 5 2

Include your tutorial lecturer's name: Simon Loop

# Part 1 - Identify a Social Problem

Use APA referencing throughout your document. [Here's a link to some explanation.](https://www.mendeley.com/guides/apa-citation-guide/)

## 1.1 Describe the Social Problem

Include the following:

- Why is this relevant?

- ...

# Part 2 - Data Sourcing

## 2.1 Load in the data

Preferably from a URL, but if not, make sure to download the data and store it in a shared location that you can load the data in from. Do not store the data in a folder you include in the Github repository!

```{r loading_data}
# gpg_data <- get_eurostat("tesem180")
# salary_data <- get_eurostat("nama_10_fte")
# saveRDS(gpg_data, "data/gpg_data")
# saveRDS(salary_data, "data/salary_data")
salary_data <- readRDS("data/salary_data")
gpg_data <- readRDS("data/gpg_data")
```

midwest is an example dataset included in the tidyverse package

## 2.2 Provide a short summary of the dataset(s)

```{r}
summary(gpg_data)

```

```{r}
summary(salary_data)

```

In this case we see 28 variables, but we miss some information on what units they are in. We also don't know anything about the year/moment in which this data has been captured.

``` r
inline_code = TRUE
```

These are things that are usually included in the metadata of the dataset. For your project, you need to provide us with the information from your metadata that we need to understand your dataset of choice.

## 2.3 Describe the type of variables included

Think of things like:

- The variables in the Salary Data Set include Average full time adjusted salary per employee per year in a given country within the EU

  - The values in the Data Set are in EUROS per year, it also includes the variable of the Time Period - 3 years in this case - 2022, 2023 and 2024. Another variable are the Countries within the EU

- Do the variables contain health information or SES information?

- Have they been measured by interviewing individuals or is the data coming from administrative sources?

*For the sake of this example, I will continue with the assignment...*

# Part 3 - Quantifying

## 3.1 Data cleaning
We perform several data cleaning steps to prepare the datasets for analysis: checking for missing values, removing incomplete observations, selecting relevant variables, and excluding aggregate regional entities.
```{r missing_values}
colSums(is.na(gpg_data))
colSums(is.na(salary_data))
```
We checked both datasets for missing values. The output shows the number of missing values in each variable.
```{r remove_missing}
gpg_data <- na.omit(gpg_data)
salary_data <- na.omit(salary_data)
```
We used na.omit() to remove observations with missing values from both datasets.
```{r select_variables}
gpg_data <- gpg_data %>%
  select(geo, TIME_PERIOD, values)
salary_data <- salary_data %>%
  filter(unit == "EUR") %>%
  select(geo, TIME_PERIOD, values)
```
We kept only the variables needed for the analysis (country, year, and values). For the salary dataset, only observations reported in euros (EUR) were retained.
```{r remove_aggregate_regions}
gpg_data <- gpg_data %>%
  filter(!geo %in% c("EA19", "EA20", "EU27_2020"))

salary_data <- salary_data %>%
  filter(!geo %in% c("EA19", "EA20", "EU27_2020"))
```
Eurostat datasets often include aggregate entities such as "EA19" (Euro area with 19 countries), "EA20" (Euro area with 20 countries), and "EU27_2020" (European Union with 27 member states, 2020 composition). We exclude these to focus exclusively on individual member states, enabling country-level comparisons.


## 3.2 Generate necessary variables

Variable 1

```{r gen_var1}
gpg_data <- gpg_data %>% rename(gpg = values)
salary_data <- salary_data %>% rename(salary = values)

merged_data <- left_join(gpg_data, salary_data, by = c("geo", "TIME_PERIOD"))

merged_data <- merged_data %>%
  mutate(gpg_severity = (gpg / 100) * salary)
cat("Merged data dimensions:", dim(merged_data), "\n")
cat("Variables:", names(merged_data), "\n")
head(merged_data)
```

Variable 2

```{r gen_var2}
# Classifying EU member states into European sub-regions
northern_eu <- c("DK", "EE", "FI", "IE", "LT", "LV", "SE")
western_eu  <- c("AT", "BE", "DE", "FR", "LU", "NL")
southern_eu <- c("CY", "ES", "GR", "HR", "IT", "MT", "PT", "SI")
eastern_eu  <- c("BG", "CZ", "HU", "PL", "RO", "SK")

region_lookup <- c(
  setNames(rep("Northern Europe", length(northern_eu)), northern_eu),
  setNames(rep("Western Europe",  length(western_eu)),  western_eu),
  setNames(rep("Southern Europe", length(southern_eu)), southern_eu),
  setNames(rep("Eastern Europe",  length(eastern_eu)),  eastern_eu)
)
merged_data <- merged_data %>%
  mutate(region = region_lookup[geo])
merged_data <- merged_data %>%
  filter(!is.na(region))
cat("Countries per region:\n")
table(merged_data$region, merged_data$TIME_PERIOD) %>% rowSums() %>% print()
```


## 3.3 Visualize temporal variation

```{r test_plot_usable}
ggplot(region_summary, aes(x = TIME_PERIOD, y = mean_gpg, color = region, group = region)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  labs(
    title = "Average Gender Pay Gap (%) over Time by European Region",
    x = "Year", y = "Mean GPG (%)", color = "Region"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## 3.4 Visualize spatial variation

```{r visualise_map}
library(maps)
library(dplyr)

world_map <- map_data("world")

code_to_name <- c(
  "AT" = "Austria",      "BE" = "Belgium",      "BG" = "Bulgaria",
  "CY" = "Cyprus",       "CZ" = "Czech Republic","DE" = "Germany",
  "DK" = "Denmark",      "EE" = "Estonia",      "ES" = "Spain",
  "FI" = "Finland",      "FR" = "France",       "GR" = "Greece",
  "HR" = "Croatia",      "HU" = "Hungary",      "IE" = "Ireland",
  "IT" = "Italy",        "LT" = "Lithuania",    "LU" = "Luxembourg",
  "LV" = "Latvia",       "MT" = "Malta",        "NL" = "Netherlands",
  "PL" = "Poland",       "PT" = "Portugal",     "RO" = "Romania",
  "SE" = "Sweden",       "SI" = "Slovenia",     "SK" = "Slovakia"
)

latest_year <- max(merged_data$TIME_PERIOD)

gpg_latest <- merged_data %>%
  filter(TIME_PERIOD == latest_year) %>%
  mutate(country_name = code_to_name[geo])

europe_map <- world_map %>%
  filter(region %in% gpg_latest$country_name)

europe_map <- europe_map %>%
  left_join(gpg_latest, by = c("region" = "country_name"))

ggplot(europe_map, aes(x = long, y = lat, group = group, fill = gpg)) +
  geom_polygon(colour = "white", linewidth = 0.2) +
  scale_fill_gradientn(
    colours = c("pink", "red", "darkred"),
    name = "GPG (%)"
  ) +
  coord_quickmap(xlim = c(-25, 45), ylim = c(35, 70)) +  
  labs(
    title = paste("Gender Pay Gap in", latest_year, "by EU Country"),
    subtitle = "Percentage difference in average earnings (women vs men)",
    caption = "Source: Eurostat (tesem180 & nama_10_fte)"
  ) +
  theme_void() +
  theme(
    plot.title   = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11)
  )

#
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations}
gpg_latest %>%
  ggplot(aes(x = region, y = gpg, fill = region)) +
  geom_boxplot(alpha = 0.8) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 3, color = "black") +
  labs(
    title = paste("Distribution of Gender Pay Gap by European Region (", latest_year, ")", sep = ""),
    subtitle = "Diamond markers indicate regional means",
    x = "European Region",
    y = "Gender Pay Gap (%)",
    fill = "Region"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 30, hjust = 1)
  )
```

```{r subpopulation_stats}
gpg_latest %>%
  group_by(region) %>%
  summarise(
    n = n(),
    mean_gpg = mean(gpg, na.rm = TRUE),
    median_gpg = median(gpg, na.rm = TRUE),
    sd_gpg = sd(gpg, na.rm = TRUE),
    min_gpg = min(gpg, na.rm = TRUE),
    max_gpg = max(gpg, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_gpg))
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}
ggplot(gpg_latest, aes(x = salary, y = gpg)) +
  geom_point(aes(color = region), size = 3, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 1) +
  geom_text(aes(label = geo), vjust = -1, size = 2.5, alpha = 0.7) +
  labs(
    title = paste("Relationship between Gender Pay Gap and Average Salary (", latest_year, ")", sep = ""),
    subtitle = "Linear regression with 95% confidence interval",
    x = "Average Annual Full-Time Adjusted Salary (EUR)",
    y = "Gender Pay Gap (%)",
    color = "Region"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

#not part of assignment
#this is wrong
#variables need to be shown with an event, event relevant for a topic
```

```{r analysis_correlation}
gpg_clean <- gpg_latest %>%
  filter(!is.na(gpg) & !is.na(salary))

# Pearson correlation test
cor_test <- cor.test(gpg_clean$salary, gpg_clean$gpg, method = "pearson")
cat("Pearson correlation coefficient (r):", round(cor_test$estimate, 3), "\n")
cat("p-value:", format(cor_test$p.value, digits = 4), "\n")
cat("95% CI: [", round(cor_test$conf.int[1], 3), ", ", round(cor_test$conf.int[2], 3), "]\n", sep = "")

# Linear regression model
lm_model <- lm(gpg ~ salary, data = gpg_clean)
cat("\nLinear Regression Summary:\n")
summary(lm_model)
```

```{r analysis_regression_table}
cat("Regression Results:\n")
cat("====================\n")
coef_summary <- summary(lm_model)$coefficients
cat(sprintf("Intercept: %.3f (SE = %.3f, p = %.4f)\n",
            coef_summary[1,1], coef_summary[1,2], coef_summary[1,4]))
cat(sprintf("Salary coefficient: %.5f (SE = %.5f, p = %.4f)\n",
            coef_summary[2,1], coef_summary[2,2], coef_summary[2,4]))
cat(sprintf("R-squared: %.3f\n", summary(lm_model)$r.squared))
cat(sprintf("Adjusted R-squared: %.3f\n", summary(lm_model)$adj.r.squared))
```

```{r analysis_region_model}
lm_extended <- lm(gpg ~ salary + region, data = gpg_clean)
cat("\nExtended Model (controlling for region):\n")
cat("========================================\n")
summary(lm_extended)
```

\`\`\`

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

Use APA referencing throughout your document.
