---
title: "Template"
author: "Studentnames and studentnumbers here"
date: "`r Sys.Date()`"
output: pdf_document
---

# Set-up your environment

```{r package_install, include=FALSE}
install.packages("tidyverse")
install.packages("yaml")
install.packages("rmarkdown")
install.packages("eurostat")
```

```{r packages}
require(tidyverse)
require(rmarkdown)
require(yaml)
require(eurostat)
```

# Title Page

Include your names: Imane Ouahi, David Abelovic, Xi Hui, Rebecca Devasia Let's do it!!! Hi Rebecca

Include the tutorial group number: 5 2

Include your tutorial lecturer's name: Simon Loop

# Part 1 - Identify a Social Problem

Use APA referencing throughout your document. [Here's a link to some explanation.](https://www.mendeley.com/guides/apa-citation-guide/)

## 1.1 Describe the Social Problem

Include the following:

- Why is this relevant?

- ...

# Part 2 - Data Sourcing

## 2.1 Load in the data

Preferably from a URL, but if not, make sure to download the data and store it in a shared location that you can load the data in from. Do not store the data in a folder you include in the Github repository!

```{r loading_data}
# gpg_data <- get_eurostat("tesem180")
# salary_data <- get_eurostat("nama_10_fte")
# saveRDS(gpg_data, "data/gpg_data")
# saveRDS(salary_data, "data/salary_data")
salary_data <- readRDS("data/salary_data")
gpg_data <- readRDS("data/gpg_data")
```

midwest is an example dataset included in the tidyverse package

## 2.2 Provide a short summary of the dataset(s)

```{r}
summary(gpg_data)

```

```{r}
summary(salary_data)

```

In this case we see 28 variables, but we miss some information on what units they are in. We also don't know anything about the year/moment in which this data has been captured.

``` r
inline_code = TRUE
```

These are things that are usually included in the metadata of the dataset. For your project, you need to provide us with the information from your metadata that we need to understand your dataset of choice.

## 2.3 Describe the type of variables included

Think of things like:

- The variables in the Salary Data Set include Average full time adjusted salary per employee per year in a given country within the EU

  - The values in the Data Set are in EUROS per year, it also includes the variable of the Time Period - 3 years in this case - 2022, 2023 and 2024. Another variable are the Countries within the EU

- Do the variables contain health information or SES information?

- Have they been measured by interviewing individuals or is the data coming from administrative sources?

*For the sake of this example, I will continue with the assignment...*

# Part 3 - Quantifying

## 3.1 Data cleaning

Say we want to include only larger distances (above 2) in our dataset, we can filter for this.

```{r missing_values}
colSums(is.na(gpg_data))
colSums(is.na(salary_data))
```

```{r remove_missing}
gpg_data <- na.omit(gpg_data)
salary_data <- na.omit(salary_data)
```

```{r select_variables}
gpg_data <- gpg_data %>%
  select(geo, TIME_PERIOD, values)
salary_data <- salary_data %>%
  filter(unit == "EUR") %>%
  select(geo, TIME_PERIOD, values)
```

```{r remove_aggregate_regions}
gpg_data <- gpg_data %>%
  filter(!geo %in% c("EA19", "EA20", "EU27_2020"))

salary_data <- salary_data %>%
  filter(!geo %in% c("EA19", "EA20", "EU27_2020"))
```

Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.

## 3.2 Generate necessary variables

Variable 1

```{r gen_var1}
gpg_data <- rename(gpg_data, gpg = values)
salary_data <- rename(salary_data, salary = values)
merged_data <- left_join(
  gpg_data,
  salary_data,
  by = c("geo", "TIME_PERIOD")
)


```

Variable 2

```{r gen_var2}
#unique(merged_data$geo)
#merged_data$region <- ifelse(
  #merged_data$geo %in% c(
    #"AT","BE","CH","DE","DK","FI",
    #"FR","IE","IS","LU","NL","NO",
    #"SE","UK"
  #),
  #"Western Europe",
  #"Eastern Europe"
#)

merged_data <- merged_data %>%
  mutate(gpg_severity = (gpg / 100) * salary)

#names(merged_data)      # lists all column names
#str(merged_data)        # shows structure and first few values
#head(merged_data)       # preview the data
```

## 3.3 Visualize temporal variation

```{r}


```

## 3.4 Visualize spatial variation

```{r visualise_map}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations}
dataset$inmetro <- dataset$inmetro %>% as.factor()
# Boxplot of poverty rate by state using the 'midwest' dataset
ggplot(dataset, aes(x = inmetro, y = percadultpoverty)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Poverty Rates by Metropolitan status (Midwest counties)",
    x = "Metropolitan Area",
    y = "Poverty Rate of Adults (%)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right"
  )
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

Use APA referencing throughout your document.
