---
title: "Template"
author: "Studentnames and studentnumbers here"
date: "`r Sys.Date()`"
output: pdf_document
---

# Set-up your environment

```{r package_install, include=FALSE}
install.packages("tidyverse")
install.packages("yaml")
install.packages("rmarkdown")
install.packages("eurostat")
install.packages("maps")
```

```{r packages}
require(tidyverse)
require(rmarkdown)
require(yaml)
require(eurostat)
require(maps)
```

# Title Page

Include your names: Imane Ouahi, David Abelovic, Xi Hui, Rebecca Devasia Let's do it!!! Hi Rebecca

Include the tutorial group number: 5 2

Include your tutorial lecturer's name: Simon Loop

# Part 1 - Identify a Social Problem

Use APA referencing throughout your document. [Here's a link to some explanation.](https://www.mendeley.com/guides/apa-citation-guide/)

## 1.1 Describe the Social Problem

Include the following:

- Why is this relevant?

- ...

# Part 2 - Data Sourcing

## 2.1 Load in the data

Preferably from a URL, but if not, make sure to download the data and store it in a shared location that you can load the data in from. Do not store the data in a folder you include in the Github repository!

```{r loading_data}
# gpg_data <- get_eurostat("tesem180")
# salary_data <- get_eurostat("nama_10_fte")
# saveRDS(gpg_data, "data/gpg_data")
# saveRDS(salary_data, "data/salary_data")
salary_data <- readRDS("data/salary_data")
gpg_data <- readRDS("data/gpg_data")
```

midwest is an example dataset included in the tidyverse package

## 2.2 Provide a short summary of the dataset(s)

```{r}
summary(gpg_data)

```

```{r}
summary(salary_data)

```

In this case we see 28 variables, but we miss some information on what units they are in. We also don't know anything about the year/moment in which this data has been captured.

``` r
inline_code = TRUE
```

These are things that are usually included in the metadata of the dataset. For your project, you need to provide us with the information from your metadata that we need to understand your dataset of choice.

## 2.3 Describe the type of variables included

Think of things like:

- The variables in the Salary Data Set include Average full time adjusted salary per employee per year in a given country within the EU

  - The values in the Data Set are in EUROS per year, it also includes the variable of the Time Period - 3 years in this case - 2022, 2023 and 2024. Another variable are the Countries within the EU

- Do the variables contain health information or SES information?

- Have they been measured by interviewing individuals or is the data coming from administrative sources?

*For the sake of this example, I will continue with the assignment...*

# Part 3 - Quantifying

## 3.1 Data cleaning

Say we want to include only larger distances (above 2) in our dataset, we can filter for this.

```{r missing_values}
colSums(is.na(gpg_data))
colSums(is.na(salary_data))
```

```{r remove_missing}
gpg_data <- na.omit(gpg_data)
salary_data <- na.omit(salary_data)
```

```{r select_variables}
gpg_data <- gpg_data %>%
  select(geo, TIME_PERIOD, values)
salary_data <- salary_data %>%
  filter(unit == "EUR") %>%
  select(geo, TIME_PERIOD, values)
```

```{r remove_aggregate_regions}
gpg_data <- gpg_data %>%
  filter(!geo %in% c("EA19", "EA20", "EU27_2020"))

salary_data <- salary_data %>%
  filter(!geo %in% c("EA19", "EA20", "EU27_2020"))
```

```{r grouping_countries}

northern_eu <- c("DK", "EE", "FI", "IE", "LT", "LV", "SE")
western_eu  <- c("AT", "BE", "DE", "FR", "LU", "NL")
southern_eu <- c("CY", "ES", "GR", "HR", "IT", "MT", "PT", "SI")
eastern_eu  <- c("BG", "CZ", "HU", "PL", "RO", "SK")

region_lookup <- c(
  setNames(rep("Northern Europe", length(northern_eu)), northern_eu),
  setNames(rep("Western Europe",  length(western_eu)),  western_eu),
  setNames(rep("Southern Europe", length(southern_eu)), southern_eu),
  setNames(rep("Eastern Europe",  length(eastern_eu)),  eastern_eu)
)

merged_data <- merged_data %>%
  mutate(region = region_lookup[geo])

# Remove unmatched countries (AL, RS, MK, ME, TR, etc.)
merged_data <- merged_data %>%
  filter(!is.na(region))

```

Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.

## 3.2 Generate necessary variables

Variable 1

```{r gen_var1}
#gpg_data <- rename(gpg_data, gpg = values)
#salary_data <- rename(salary_data, salary = values)
#merged_data <- left_join(
  #gpg_data,
  #salary_data,
  #by = c("geo", "TIME_PERIOD")
#)

region_summary <- merged_data %>%
  group_by(region, TIME_PERIOD) %>%
  summarise(
    mean_gpg = mean(gpg, na.rm = TRUE),
    mean_severity = mean(gpg_severity, na.rm = TRUE),
    .groups = "drop"
  )

```

Variable 2

```{r gen_var2}
#unique(merged_data$geo)
#merged_data$region <- ifelse(
  #merged_data$geo %in% c(
    #"AT","BE","CH","DE","DK","FI",
    #"FR","IE","IS","LU","NL","NO",
    #"SE","UK"
  #),
  #"Western Europe",
  #"Eastern Europe"
#)

merged_data <- merged_data %>%
  mutate(gpg_severity = (gpg / 100) * salary)
```

## 3.3 Visualize temporal variation


```{r test plot 2}
ggplot(merged_data, aes(x = TIME_PERIOD, y = gpg, group = geo)) +
  geom_line(color = "steelblue") +
  geom_point(color = "steelblue") +
  facet_wrap(~ geo, scales = "free_y") +
  labs(
    title = "Gender Pay Gap over Time (by country)",
    x = "Year",
    y = "GPG (%)"
  ) +
  theme_minimal()
```
```{r test plot 3}

```

```{r test_plot_usable}
ggplot(region_summary, aes(x = TIME_PERIOD, y = mean_gpg, color = region, group = region)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  labs(
    title = "Average Gender Pay Gap (%) over Time by European Region",
    x = "Year", y = "Mean GPG (%)", color = "Region"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```


## 3.4 Visualize spatial variation

```{r visualise_map}
library(maps)
library(dplyr)

# 1. Get the world map data as a data frame
world_map <- map_data("world")

# 2. Create a lookup from your country codes (geo) to full English country names
#    Only include the EU countries you kept after filtering.
code_to_name <- c(
  "AT" = "Austria",      "BE" = "Belgium",      "BG" = "Bulgaria",
  "CY" = "Cyprus",       "CZ" = "Czech Republic","DE" = "Germany",
  "DK" = "Denmark",      "EE" = "Estonia",      "ES" = "Spain",
  "FI" = "Finland",      "FR" = "France",       "GR" = "Greece",
  "HR" = "Croatia",      "HU" = "Hungary",      "IE" = "Ireland",
  "IT" = "Italy",        "LT" = "Lithuania",    "LU" = "Luxembourg",
  "LV" = "Latvia",       "MT" = "Malta",        "NL" = "Netherlands",
  "PL" = "Poland",       "PT" = "Portugal",     "RO" = "Romania",
  "SE" = "Sweden",       "SI" = "Slovenia",     "SK" = "Slovakia"
)

# 3. Filter merged_data to the latest year and add the full country name
latest_year <- max(merged_data$TIME_PERIOD)

gpg_latest <- merged_data %>%
  filter(TIME_PERIOD == latest_year) %>%
  mutate(country_name = code_to_name[geo])

# 4. Filter the world map to only the European countries in your data
europe_map <- world_map %>%
  filter(region %in% gpg_latest$country_name)

# 5. Join the GPG data to the map data
europe_map <- europe_map %>%
  left_join(gpg_latest, by = c("region" = "country_name"))

# 6. Draw the choropleth map
ggplot(europe_map, aes(x = long, y = lat, group = group, fill = gpg)) +
  geom_polygon(colour = "white", linewidth = 0.2) +
  scale_fill_gradientn(
    colours = c("darkgreen", "yellow", "red"),
    name = "GPG (%)"
  ) +
  coord_quickmap(xlim = c(-25, 45), ylim = c(35, 70)) +  # zoom to Europe
  labs(
    title = paste("Gender Pay Gap in", latest_year, "by EU Country"),
    subtitle = "Percentage difference in average earnings (women vs men)",
    caption = "Source: Eurostat (tesem180 & nama_10_fte)"
  ) +
  theme_void() +
  theme(
    plot.title   = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11)
  )
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations}
dataset$inmetro <- dataset$inmetro %>% as.factor()
# Boxplot of poverty rate by state using the 'midwest' dataset
ggplot(dataset, aes(x = inmetro, y = percadultpoverty)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Poverty Rates by Metropolitan status (Midwest counties)",
    x = "Metropolitan Area",
    y = "Poverty Rate of Adults (%)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right"
  )
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

Use APA referencing throughout your document.


